---
title: "Project II: Cuckoo Eggs Dimension"
author: "Nangamso Lindeni"
format: pdf
toc: true
toc-depth: 3
toc-title: "Contents"
editor: visual
---

## Introduction

The Common Cuckoo (*Cuculus canorus*) is a bird species known for its distinctive behaviours and characteristics, particularly its unusual reproductive strategy. As brood parasites, cuckoos lay their eggs in the nests of other, often smaller, bird species. The host bird then incubates the cuckoo egg and unknowingly raises the chick as its own (Davies, 2011; Lahti, 2005). According to Kruger and Davies (2003), Davies (2011), and Lahti (2005), the survival of the cuckoo species hinges on their remarkable adaptation strategies, which include evading host defences through deception and matching their egg and chick development with the host's incubation and provisioning patterns.

Research conducted in Russia has provided evidence of evolutionary adaptation in cuckoos, particularly in their ability to vary egg size to match that of their hosts. This adaptation reduces the likelihood of egg rejection, as host birds are more inclined to reject eggs that are significantly different or larger than their own clutch (Meshcheryagina et al., 2018). The study found that while the breadth of cuckoo eggs closely matched that of the host species, the length did not, suggesting a trade-off that limits the cuckoo's ability to maintain larger egg volumes. This alignment in egg breadth is likely an evolved response to avoid host rejection (Meshcheryagina *et al*., 2018).

This project aims to investigate the egg dimensions—length and breadth—of six different cuckoo bird species, analysing 120 eggs. Various statistical analyses will be performed to determine differences in these dimensions both between and within the species.

## Importing Data

For data manipulation, visualisation and analysis and to effectively read and access the relevant cuckoo data, the "tidyverse" was installed and activated. The read.csv () function was used to import and read the data that is in CSV format. The data was assigned a new name termed cuckoo_data.

```{r}
library(tidyverse)

cuckoo_data <- read.csv("https://raw.githubusercontent.com/xrander/bootcamp-test/master/data/cuckoo.csv")

```

To view and understand the contents of the data names() function was used to return the names of the columns in the cuckoo_data, this also ensures that the data has been imported correctly. The names of the the cuckoo_data are "length" "breadth" "species"

```{r}
names(cuckoo_data)
```

The head() function was implemented to preview the first 10 rows of the data, while the tail() function was run to preview the last 10 rows of the dataset, thus making it easier to observe patterns in the data.

```{r}
head(cuckoo_data, n = 10)
tail(cuckoo_data, n = 10)
```

To further understand and observe the cuckoo_data the car::some() function was run, first setting a seed, to randomly select and generate a sample of rows from the cuckoo_data dataset. The set. seed() allows the same results every time the code is run.

```{r}
set.seed(3891)
car::some(cuckoo_data)
```

## Exploratory Data Analysis

To investigate the main characteristics and understand the structure of the dataset, various analyses were conducted. The summary() function was run to return a key statistics overview of the data which include minimum, maximum, mean, median and quartiles for numerical variables of the data. The str() function gave statistics of the three variables in the data frame; length, breadth, species

```{r}
summary(cuckoo_data)
```

The str() function was run to show the structure of the data along with the type and length of variables in the dataset. For cuckoo_data the str() function showed that the data has 120 observations, 3 variables, variables length and breadth are numerical, while species variable is character.

```{r}
str(cuckoo_data)
```

To identify distinct species present in the data frame, the unique() function was used and it generated six species names in the data, Meadow Pipit, Tree Pipit, Hedge Sparrow, Robin, Pied Wagtail and Wren.

```{r}
unique(cuckoo_data$species)
```

To check if the dataset had any missing values, the `is.na()` function was used, which returned `FALSE` for all columns, indicating no missing values. To obtain a more concise and straightforward summary, the `colSums()` function was applied to sum up the `FALSE` values for each column, resulting in a count of `0` for all columns, thereby confirming that there was no missing data in the dataset.

```{r}
colSums(is.na(cuckoo_data))
```

The skim() function was run to obtain an extended structure of the data and gave statical values of the data such as mean, median standard deviation along with missing values and histograms of the data . This function used along with skimr package combined the functionality if the summary() and str() function.

```{r}
skimr::skim(cuckoo_data)
```

## Visualisation

```{r}
library(ggplot2)
ggplot(cuckoo_data, aes(species, breadth)) + 
  geom_boxplot() + 
  labs(x = "Species",
       y = "Breadth (mm)", 
       title = "Comparing Breadth size between Species")

ggplot(cuckoo_data, aes(species, length)) +
  geom_boxplot() +
  labs(x = "Species",
       y = "Length (mm)", 
       title = "Comparing Length size bewteen Species")


ggplot(cuckoo_data, aes(x = length, y = breadth, color = species)) +
  geom_point() +
  labs(x = "Egg Length (mm)",
       y = "Egg Breadth (mm)", 
       title = "Scatter Plot of Egg Length vs Breadth by Species")

ggplot(cuckoo_data, aes(x = length, fill = species)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot of Egg Length by Species", x = "Egg Length (mm)", y = "Density")

ggplot(cuckoo_data, aes(x = breadth, fill = species)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot of Egg Breadth by Species", x = "Egg Breadth (mm)", y = "Density")


```

## Inferential Statistical Analysis

Two-way ANOVA to test how the species (dependent variable) influences the length (independent variable) of eggs from different bird species (independent variable).

```{r}
aov(breadth ~ species, data = cuckoo_data)
```

Two-way ANOVA test to test how bird species independently influence breadth.

```{r}
aov(length ~ species, data = cuckoo_data)

```

Multivariate Linear Regression to assess possible linear relationship of length and breadth measurements of eggs based on species.

```{r}
library(ggplot2)

```

## Discussion and Conclusion
