---
title: "Project II: Cuckoo Eggs Dimension"
author: "Nangamso Lindeni"
format: pdf
toc: true
toc-depth: 3
toc-title: "Contents"
editor: visual
---

## Introduction

The Common Cuckoo (*Cuculus canorus*) is a bird species known for its distinctive behaviours and characteristics, particularly its unusual reproductive strategy. As brood parasites, cuckoos lay their eggs in the nests of other, often smaller, bird species. The host bird then incubates the cuckoo egg and unknowingly raises the chick as its own (Davies, 2011; Lahti, 2005). According to Kruger and Davies (2003), Davies (2011), and Lahti (2005), the survival of the cuckoo species hinges on their remarkable adaptation strategies, which include evading host defences through deception and matching their egg and chick development with the host's incubation and provisioning patterns.

Research conducted in Russia has provided evidence of evolutionary adaptation in cuckoos, particularly in their ability to vary egg size to match that of their hosts. This adaptation reduces the likelihood of egg rejection, as host birds are more inclined to reject eggs that are significantly different or larger than their own clutch (Meshcheryagina et al., 2018). The study found that while the breadth of cuckoo eggs closely matched that of the host species, the length did not, suggesting a trade-off that limits the cuckoo's ability to maintain larger egg volumes. This alignment in egg breadth is likely an evolved response to avoid host rejection (Meshcheryagina *et al*., 2018).

This project aims to investigate the egg dimensions—length and breadth—of six different cuckoo bird species, analysing 120 eggs. Various statistical analyses will be performed to determine differences in these dimensions both between and within the species.

## Importing Data

For data manipulation, visualisation and analysis and to effectively read and access the relevant cuckoo data, the "tidyverse" and "readr" packages were installed and activated. The read.csv () function was used to import and read the data that is in CSV format. The data was assigned a new name termed cuckoo_data.

```{r}
library(tidyverse)
library(readr)

cuckoo_data <- read.csv("https://raw.githubusercontent.com/xrander/bootcamp-test/master/data/cuckoo.csv")

```

To view and understand the contents of the data names() function was used to return the names of the columns in the cuckoo_data, this also ensures that the data has been imported correctly.

```{r}
names(cuckoo_data)
```

The head() function was implemented to preview the first 10 rows of the data, while the tail() function was run to preview the last 10 rows of the dataset, thereby making it easier to observe patterns in the data.

```{r}
head(cuckoo_data, n = 10)
tail(cuckoo_data, n = 10)
```

Lastly, to further understand and observe the cuckoo_data the car::some() function was run, first setting a seed, to randomly select and generate a sample of rows from cuckoo_data dataset. The set.seed() allows the same results everytime the code is run.

```{r}
set.seed(3891)
car::some(cuckoo_data)
```

## Exploratory Data Analysis

To investigate the main characteristics of the data

```{r}
summary(cuckoo_data)

str(cuckoo_data)

unique(cuckoo_data$species)

colSums(is.na(cuckoo_data))

skimr::skim(cuckoo_data)

```

## Visualisation

```{r}
library(ggplot2)
ggplot(cuckoo_data, aes(species, breadth)) + 
  geom_boxplot() + 
  labs(x = "Species",
       y = "Breadth (mm)", 
       title = "Comparing Breadth size between Species")

ggplot(cuckoo_data, aes(species, length)) +
  geom_boxplot() +
  labs(x = "Species",
       y = "Length (mm)", 
       title = "Comparing Length size bewteen Species")


ggplot(cuckoo_data, aes(x = length, y = breadth, color = species)) +
  geom_point() +
  labs(x = "Egg Length (mm)",
       y = "Egg Breadth (mm)", 
       title = "Scatter Plot of Egg Length vs Breadth by Species")

ggplot(cuckoo_data, aes(x = length, fill = species)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot of Egg Length by Species", x = "Egg Length (mm)", y = "Density")

ggplot(cuckoo_data, aes(x = breadth, fill = species)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot of Egg Breadth by Species", x = "Egg Breadth (mm)", y = "Density")

cuckoo_data2 <- gather(cuckoo_data, key = "measurement", value = "value", length, breadth)

ggplot(cuckoo_data2, aes(x = species, y = value, fill = measurement)) +
  geom_bar(stat = "identity") +
  labs(title = "Stacked Bar Graph of Egg Length and Breadth by Species",
       x = "Species",
       y = "Measurement (mm)",
       fill = "Measurement") +
  theme_minimal()
```

## Inferential Statistical Analysis

Two-way ANOVA to test how the species (dependent variable) influences the length (independent variable) of eggs from different bird species (independent variable).

```{r}
aov(breadth ~ species, data = cuckoo_data)
```

Two-way ANOVA test to test how bird species independently influence breadth.

```{r}
anova_length <- aov(length ~ species, data = cuckoo_data)
summary(anova_length)
```

Multivariate Linear Regression to assess possible linear relationship of length and breadth measurements of eggs based on species.

```{r}
library(ggplot2)

```

## Discussion and Conclusion
